---
title: "Practical Machine Learning Course Project"
author: "Sushant  Pagnis"
date: "27 May 2018"
output: html_document
---


## Intro

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.
In this project, our goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. The goal of this project is to predict the manner in which they did the exercise.

## Loading and Cleaning the data

Doing the prerequisites by loading the libraries

```{r}
library("caret")

if(!file.exists("pml-training.csv")){download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile = "pml-training.csv")}

if(!file.exists("pml-testing.csv")){download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile = "pml-testing.csv")}

trainingData<- read.csv("pml-training.csv", sep=",", header=TRUE, na.strings = c("NA","",'#DIV/0!'))
testingData<- read.csv("pml-testing.csv", sep=",", header=TRUE, na.strings = c("NA","",'#DIV/0!'))

dim(trainingData)

dim(testingData)

trainingData <- trainingData[,(colSums(is.na(trainingData)) == 0)]
dim(trainingData)

testingData <- testingData[,(colSums(is.na(testingData)) == 0)]
dim(testingData)
```

## Preprocessing the data

We also remove the values tending to zero as they will have minimal impact

```{r}
numericalsIdx <- which(lapply(trainingData, class) %in% "numeric")

preprocessModel <-preProcess(trainingData[,numericalsIdx],method=c('knnImpute', 'center', 'scale'))
pre_trainingData <- predict(preprocessModel, trainingData[,numericalsIdx])
pre_trainingData$classe <- trainingData$classe

pre_testingData <-predict(preprocessModel,testingData[,numericalsIdx])

nzv <- nearZeroVar(pre_trainingData,saveMetrics=TRUE)
pre_trainingData <- pre_trainingData[,nzv$nzv==FALSE]

nzv <- nearZeroVar(pre_testingData,saveMetrics=TRUE)
pre_testingData <- pre_testingData[,nzv$nzv==FALSE]
```


## Validation Set

We want a 75% observation training dataset to train our model. We will then validate it on the last 70%

```{r}
set.seed(100)
idxTrain<- createDataPartition(pre_trainingData$classe, p=3/4, list=FALSE)
training<- pre_trainingData[idxTrain, ]
validation <- pre_trainingData[-idxTrain, ]
dim(training) ; dim(validation)
```

## Train the model

We train a model using random forest with a cross validation of 5 folds to avoid overfitting.

```{r}
library(randomForest)
modFitrf <- train(classe ~., method="rf", data=training, trControl=trainControl(method='cv'), number=5, allowParallel=TRUE, importance=TRUE )
modFitrf
```

## Interpretation

```{r}
varImpPlot(modFitrf$finalModel, sort = TRUE, type = 1, pch = 19, col = 1, cex = 0.6, main = "Importance of the Individual Principal Components")
```

This plot shows each of the principal components in order from most important to least important.

## Cross Validation Testing and Out-of-Sample Error Estimate

Let’s apply our training model on our testing database, to check its accuracy.

# Accuracy and Estimated out of sample error

```{r}
predValidRF <- predict(modFitrf, validation)
confus <- confusionMatrix(validation$classe, predValidRF)
confus$table
```

We can notice that there are very few variables out of this model.

```{r}
accuracy <- postResample(validation$classe, predValidRF)
modAccuracy <- accuracy[[1]]
modAccuracy

out_of_sample_error <- 1 - modAccuracy
out_of_sample_error
```

The estimated accuracy of the model is 99.7% and the estimated out-of-sample error based on our fitted model applied to the cross validation dataset is 0.3%.

## Application of this model on the 20 test cases provided

We have already cleaned the test data base. We delete the “problem id” column as it is useless for our analysis.
Here are our results.

```{r}
pred_final <- predict(modFitrf, pre_testingData)
pred_final
```


